Kindly Bent to Free Us

### 2019-09-19 kindly/popl20-rebuttal.md

Thanks to the reviewers for their thoughtful comments.

# Reviewer A

> How is proper nesting of regions enforced, for both the type system and region inference in Appendix A?  And for the example I asked about above ( $\{|\ldots\{|\ldots|\}^2|\}^3$ )? Or does the paper simply give a nesting intuition that's not necessary because the indices don't actually matter for soundness?

We assume that the type checker operates on a program where
region annotations are inserted by the automatic
annotation phase provided in Appendix A.
So our typing rules do not check the nesting because they are supposed
to work only on programs with well-nested regions (a similar situation
occurs in type-based program analysis, where the analysis assumes a
well-typed program and only reasons about additional properties on values).
Unfortunately, the annotation rules have a typo: The recursive
call in "Rewrite-Region" should be on $n+1$.
Insertion of annotations is completely deterministic (which solves principality concerns)
and ensures that indices grow as regions are nested (which rules out all the given
problematic examples).
This is sufficient since there is no way to hide the level of a borrow
and smuggle it past a region boundary. 
If that was the case, for instance with existential types, we would require a
more sophisticated region system with abstract variables, as suggested
by the reviewer.

We agree that our assumption should be made clearer and will improve
the respective  explanations.

> Please explain why/whether Theorem F.2's statement properly enforces linearity

Peter??

# Reviewer B

## Modules

Affe's usage of subsumption also interacts with type abstraction and modules
For instance, the `File.t` type presented in Figure 1a could be implemented
as a file descriptor (i.e., an integer), which is unrestricted. Subsumption
then allow the unrestricted type to be exposed as affine.
Such mechanism allow programmers to implement their own primitives on
abstract types.
As usual with type-level programming, they must also ensure
the soundness of such primitives
Modules and subsumption are used similarly in Alms, and allow
to emulate the `unsafe` construct in Rust.

# Reviewer C

## Simplification and subkinding

The reviewer ignores Affe's use of subsumption which allows us to
simplify many type signatures without loosing expressiveness.
In particular, as highlighted in section 1.1, the compose function

    let compose f g x = f (g x)
    # compose : (k <= k1) => (b -k-> c) -> (a -k1-> b) -k-> (a -k1-> c)

admits having `f` linear and `g` affine with a linear outcome:
Because affine is a subkind of linear, `g`'s multiplicity can be subsumed to
linear, both k and k1 are instantiated to linear, and the resulting
function type is annotated as linear. This approach follows the
treatment of structural subtyping in HM(X).

## Regions and borrowing

The session types interface that we present is using a functional
style with linear types to demonstrate that this application is in
scope of our system and that  Affe's type inference
handles it gracefully, without extra annotations.

Borrows are more useful for imperative
programming and we are not claiming to apply them to session types.
As explained in Section 2.3, the goal of Affe is to support both
styles, allowing the programmer to choose the most appropriate one for
the task at hand.

Dealing with borrows for session-typed channels would require a major
extension with typestates, which is notoriously incompatible with type
inference. 
However, one could contemplate operations to extract channel metadata
(IP , bytes sent, etc) from a  shared borrow of a channel as in 
`get_address: &(_ channel) -> ip_address`, 
which is typable in Affe.
As the main reference to the channel cannot be used while a borrow is active, such
operations would be sound.

## Formalization

We disagree that the formalization is incomplete.
In particular, the handling of the range of borrows is detailed in Section 3.2.4 on regions.
Furthermore, the lack of deconstruction form for borrows of abstract types is
intended: only the appropriate primitives can act on borrows, because the semantics
of borrow is specific to a given abstract type: For instance the
shared borrow in `read_line : &file -> string` may lead to data races,
whereas `&!file -> string` doesn't.

For space reasons and because the current article focuses on linearity, borrowing,
and inference, we decided not to detail the handling of matching on pairs.
We agree that such aspect is important, and we will expand on this
topic if the article is accepted.

> Related work

????

# Reviewer D

> Partial inference and GADT

Even if a language does not admit full type inference, we believe, 
there should be a well-defined core language which does.
This way, a programmer who steps over the boundary knows when type
annotations are needed.
In particular, it is possible (and desirable) to extend Affe with existential types
by requiring type annotations, similarly to Alms.
The results we present nevertheless garantee that
full type inference is supported in the absence of existentials.

### 2019-09-19 kindly/popl20-rebuttal.md

Thanks to the reviewers for their thoughtful comments. Were the paper to be accepted, we would take their comments into account and revise our article in the following manner:
- Streamline the description of examples in Section 2 to focus on Affe's novelties
- Present automatic region annotation in the body of the article, and clarifies its role in the type system.
- Present the pattern matching rules, both typing and semantic.
- Improve the explanation through new examples for the semantics and the type inference (an example for constraint solving is already present in the appendix).

@@

# Reviewer A

> How is proper nesting of regions enforced, for both the type system and region inference in Appendix A?  And for the example I asked about above ( $\{|\ldots\{|\ldots|\}^2|\}^3$ )? Or does the paper simply give a nesting intuition that's not necessary because the indices don't actually matter for soundness?

We assume that the type checker operates on a program where
region annotations are inserted by the automatic
annotation phase provided in Appendix A.
So our typing rules do not check the nesting because they are supposed
to work only on programs with well-nested regions (a similar situation
occurs in type-based program analysis, where the analysis assumes a
well-typed program and only reasons about additional properties on values).
Unfortunately, the annotation rules have a typo: The recursive
call in "Rewrite-Region" should be on $n+1$.
Insertion of annotations is completely deterministic (which solves principality concerns)
and ensures that indices grow as regions are nested (which rules out all the given
problematic examples).
This is sufficient since there is no way to hide the level of a borrow
and smuggle it past a region boundary. 
If that was the case, for instance with existential types, we would require a
more sophisticated region system with abstract variables, as suggested
by the reviewer.

We agree that our assumption should be made clearer and will improve
the associated explanations.

@@

Peter??

# Reviewer B

## Modules

Affe's usage of subsumption also interacts with type abstraction and modules
For instance, the `File.t` type presented in Figure 1a could be implemented
as a file descriptor (i.e., an integer), which is unrestricted. Subsumption
then allow the unrestricted type to be exposed as affine.
Such mechanism allow programmers to implement their own primitives on
abstract types. Programmers are then in charge of ensuring soundness of such primitives. This feature, which is also present in Alms, allows to partially emulate Rust's `unsafe` construct.

### 2019-09-19 kindly/popl20-rebuttal.md

Thanks to the reviewers for their thoughtful comments. 
Were the paper to be accepted, we would take their comments into account and revise our article in the following manner:
- Streamline the description of examples in Section 2 to focus on Affe's novelties
- Present automatic region annotation in the body of the article and clarify its role in the type system.
- Present the pattern matching rules, both typing and semantic.
- Improve the explanation through new examples for the semantics and the type inference (an example for constraint solving is already present in the appendix).

@@

We assume that the type checker operates on a program where
region annotations are inserted by the automatic
annotation phase provided in Appendix A.
So our typing rules do not check the nesting because they are supposed
to work only on programs with well-nested regions (a similar situation
occurs in type-based program analysis, where the analysis assumes a
well-typed program and only reasons about additional properties on values).
Unfortunately, the annotation rules have a typo: The recursive
call in "Rewrite-Region" should be on $n+1$.
Insertion of annotations is completely deterministic (which solves principality concerns)
and ensures that indices grow as regions are nested (which rules out all the given
problematic examples).
This is sufficient since there is no way to hide the level of a borrow
and smuggle it past a region boundary. 
If that was the case, for instance with existential types, we would require a
more sophisticated region system with abstract variables, as suggested
by the reviewer.

We agree that our assumption should be made clearer and will improve
the associated explanations.

> Please explain why/whether Theorem F.2's statement properly enforces linearity

We don't understand the construction outlined by the reviewer. We
believe that the situation cannot arise because the
different parts of the environment are always disjoint. The
Region/SRegion rule moves a binding from the resource part to a borrow
part (mutable or immutable). This part of the formalization is
unfortunately a bit opaque as it is hidden in the environment
formation (L 2113-2123).
The Region typing rule transforms the typing context (for example, L
2375); the corresponding transformation of the run-time environment is
left implicit.

We will try to improve the writeup for this issue in the revision.

> The proof of F.2 seems to omit the cases of the semantics that might
> modify the heap

We put the cases that we felt were most important.

> I cannot find type rules for Create/Observe/Destroy in the supplementary material (my apologies if I missed them somewhere).

That's an unfortunate editing mistake.

@@

We disagree that the formalization is incomplete.
In particular, the handling of the range of borrows is detailed in Section 3.2.4 on regions.
Furthermore, the lack of deconstruction form for borrows of abstract types is
intended: only the appropriate primitives can act on borrows, because the semantics
of borrow is specific to a given abstract type: For instance the
shared borrow in `read_line : &file -> string` may lead to data races,
whereas `&!file -> string` doesn't.

For space reasons and because the current article focuses on linearity, borrowing,
and inference, we decided not to detail the handling of matching on pairs.
We agree that such aspect is important, and we will expand on this
topic if the article is accepted.

> Related work

Correct, Quill comes with a usage-counting semantics. What we meant to
say is that their semantics isn't store based.

Alms is a seminal work that supports many interesting examples, but it
comes with a heavy price in terms of complexity (e.g., dependent
kinds, lub operator in types) and it is not suitable for type
inference. Most of the examples in the Alms paper exploit the
genericity of existentials to, e.g., connect resources with
capabilities. As Affe doesn't have unrestricted existentials, the
corresponding examples are unlikely to work in Affe as they are stated
in the Alms paper.

The examples in section 2 are not aimed at comparing with Alms, rather
we want to demonstrate that the idea put forward in the Alms paper (to
annotate abstract types in module signatures) also works with Affe's
type inference.

We will insert a citation of the Gan, Tov, and Morrisett paper with
the copy-on-write array.

### 2019-09-20 kindly/popl20-rebuttal.md

We thanks the reviewers for their thoughtful comments. We will take their comments into account and revise our article in the following manner:
- Streamline the description of examples in Section 2 to focus on Affe's novelties
- Present automatic region annotation in the body of the article and clarify its role in the type system.
- Present the pattern matching rules, both typing and semantic.
- Improve the explanation through new examples for the semantics and the type inference (an example for constraint solving is already present in the appendix).

@@

That's an unfortunate editing mistake.

# Reviewer B

## Modules

Affe's usage of subsumption also interacts with type abstraction and modules
For instance, the `File.t` type presented in Section 1 could be implemented
as a file descriptor (i.e., an integer), which is unrestricted. Subsumption
then allow the unrestricted type to be exposed as affine.
Such mechanism allow programmers to implement their own primitives on
abstract types. Programmers are then in charge of ensuring soundness of such primitives. This feature, which is also present in Alms, allows to partially emulate Rust's `unsafe` construct.

@@

For space reasons and because the current article focuses on linearity, borrowing,
and inference, we decided not to detail the handling of matching on pairs.
We agree that such aspect is important, and we will expand on this
topic if the article is accepted.

> Related work

Correct, Quill comes with a usage-counting semantics. What we meant to
say is that their semantics isn't store based.

Alms is a seminal work that supports many interesting examples, but it
comes with a heavy price in terms of complexity (e.g., dependent
kinds, lub operator in types) and it is not suitable for type
inference. Most of the examples in the Alms paper exploit the
genericity of existentials to, e.g., connect resources with
capabilities. In order to provide full type inference
Affe doesn't have unrestricted existentials, and the
corresponding examples are unlikely to work in Affe as they are stated
in the Alms paper.

@@

# Reviewer D

> Partial inference and GADT

Even if a language does not admit full type inference, we believe, 
there should be a well-defined core language which does.
This way, a programmer who steps over the boundary knows when type
annotations are needed.
In particular, it is possible (and desirable) to extend Affe with existential types
by requiring type annotations, similarly to Alms.
The results we present nevertheless garantee that
full type inference is supported in the absence of existentials.

> the paper would strongly benefit from more examples that explain why the type inference algorithm works the way it does.

One such example is available for constraint solving in the Appendix B.1, but we agree
that it should be extended, and part of the main body.

### 2019-09-20 kindly/popl20-rebuttal.md

We thank the reviewers for their thoughtful comments.
We will take their comments into account and revise our article as follows.
* Streamline the description of examples in Section 2 to focus on Affe's novelties
* Present automatic region annotation in the body of the article and clarify its role in the type system.
* Present the pattern matching rules, both typing and semantic.
* Improve the explanation through new examples for the semantics and
  the type inference (the appendix contains an example for constraint solving).

@@

# Reviewer A

> How is proper nesting of regions enforced, for both the type system and region inference in Appendix A?  And for the example I asked about above ( $\{|\ldots\{|\ldots|\}^2|\}^3$ )? Or does the paper simply give a nesting intuition that's not necessary because the indices don't actually matter for soundness?

We assume that the type checker operates on a program where
region annotations are inserted by the
annotation algorithm provided in Appendix A.
Hence, our typing rules do not check the nesting because they are supposed
to work only on programs with well-nested regions (a similar situation
occurs in type-based program analysis, where the analysis assumes a
well-typed program and only reasons about additional properties on values).
Unfortunately, the annotation rules have a typo: The recursive
call in "Rewrite-Region" should be on $n+1$.
Insertion of annotations is deterministic (which solves principality concerns)
and ensures that indices grow as regions are nested (which rules out all the
problematic examples).
This is sufficient because there is no way to hide the level of a borrow
and smuggle it past a region boundary. 
If that was the case, for instance with existential types, we would require a
more sophisticated region system with abstract variables, as suggested
by the reviewer.

@@

We agree that our assumption should be made clearer and will improve
the associated explanations.

> Please explain why/whether Theorem F.2's statement properly enforces linearity

As far as we understand the construction outlined by the reviewer, we
believe that the situation cannot arise because the
different parts of the environment are always disjoint. This is
ensured at a high level by splitting (in the proof case for application). The
Region/SRegion rule moves a binding from the resource part to a borrow
part (mutable or immutable). This part of the formalization is
unfortunately a bit opaque as it is hidden in the environment
formation (L 2113-2123).
The Region typing rule transforms the typing context (for example, L
2375); the corresponding transformation of the run-time environment is
left implicit.

We will improve the writeup for this issue in the revision.

@@

That's an unfortunate editing mistake.

# Reviewer B

## Modules

Affe's usage of subsumption interacts with type abstraction and modules.
For instance, the `File.t` type presented in Section 1 could be implemented
as a file descriptor (i.e., an integer), which is unrestricted. Subsumption
allow us to expose the unrestricted type as an affine abstract type.
Such a mechanism allows programmers to implement their own primitives on
abstract types. Programmers are then in charge of ensuring soundness
of such primitives. This feature, which is also present in Alms,
can emulate Rust's `unsafe` construct in many cases.

@@

# Reviewer C

## Simplification and subkinding

The reviewer overlooks the presence of kind subsumption in Affe, which
allows us to simplify type signatures without loosing
expressiveness.  In particular, as highlighted in section 1.1, the
compose function

@@

    let compose f g x = f (g x)
    # compose : (k <= k1) => (b -k-> c) -> (a -k1-> b) -k-> (a -k1-> c)

admits having `f` linear and `g` affine with a linear outcome:
Because affine is a subkind of linear, `g`'s multiplicity can be subsumed to
linear, both k and k1 are instantiated to linear, and the resulting
function type is annotated as linear. This approach follows the
treatment of structural subtyping in HM(X).

## Regions and borrowing

The session type interface that we present is using a functional
style with linear types to demonstrate that this application is in
scope of our system and that Affe's type inference
handles it gracefully, without extra annotations.

Borrows are primarily useful for imperative
programming and we are not claiming to apply them to session types.
As explained in Section 2.3, the goal of Affe is to support both
styles, allowing the programmer to choose the most appropriate one for
the task at hand.

@@

Dealing with borrows for session-typed channels would require a major
extension with typestates, which is notoriously incompatible with type
inference. 
However, one could contemplate operations to extract channel metadata
(IP , bytes sent, etc) from a shared borrow of a channel as in 
`get_address: &(_ channel) -> ip_address`, 
which is typable in Affe.
As the main reference to the channel cannot be used while a borrow is active, such
operations would be sound.

## Formalization

We disagree that the formalization is incomplete.
In particular, the handling of the range of borrows is detailed in Section 3.2.4 on regions.
Furthermore, the lack of deconstruction form for borrows of abstract types is
intended: only the appropriate primitives can act on borrows, because the semantics
of borrow is specific to a given abstract type: For instance the
shared borrow in `read_line : &file -> string` may lead to data races,
whereas `&!file -> string` doesn't.

For space reasons and because the current article focuses on linearity, borrowing,
and inference, we decided not to detail the handling of matching on pairs.
We agree that this aspect is important and we will expand on this
topic in future revisions.

@@

> Related work

Correct, Quill comes with a usage-counting semantics. What we meant to
say is that their semantics isn't store based.

Alms is a seminal work that supports many interesting examples, but it
comes with a heavy price in terms of complexity (e.g., dependent
kinds, lub operator in types) and it is not suitable for type
inference. Most of the examples in the Alms paper exploit the
genericity of existentials to, e.g., connect resources with
capabilities. 
Affe doesn't have unrestricted existentials so as to support full type
inference. The corresponding examples are unlikely to work in Affe as
they are stated in the Alms paper.

@@

The examples in section 2 are not aimed at comparing with Alms, rather
we want to demonstrate that the idea put forward in the Alms paper (to
annotate abstract types in module signatures) also works with Affe's
type inference.

We will insert a citation of the Gan, Tov, and Morrisett paper along with
the copy-on-write array example.

@@

# Reviewer D

> Partial inference and GADT

Even if a language does not admit full type inference, we believe
there should be a well-defined core language which does.
This way, a programmer who steps over the boundary knows when type
annotations are needed.
In particular, it is possible (and desirable) to extend Affe with existential types
by requiring type annotations, similarly to Alms.
The results we present nevertheless garantee that
full type inference is supported in the absence of existentials.

> the paper would strongly benefit from more examples that explain why the type inference algorithm works the way it does.

One such example is available for constraint solving in the Appendix B.1, but we agree
that it should be extended and part of the main body.

### 2019-09-20 kindly/popl20-rebuttal.md

> Related work

Correct, Quill comes with a usage-counting semantics. What we meant to
say is that their semantics isn't store based.

Alms is a seminal work that supports many interesting examples, but it
comes with a heavy price in terms of complexity (e.g., dependent
kinds, lub operator in types) and it is not suitable for type
inference. Most of the examples in the Alms paper exploit the
genericity of existentials to, e.g., connect resources with
capabilities. 
Affe doesn't have unrestricted existentials so as to support full type
inference. The corresponding examples are unlikely to work in Affe as
they are stated in the Alms paper.

The examples in section 2 are not aimed at comparing with Alms, rather
we want to demonstrate that Alm's use of modules is compatible with Affe's inference, as highlighted in Reviewer B's answer.

### 2019-10-25 kindly/infer/infer.tex
Improve placement of figures.
Fixed typos from reviews and own review.

An important contribution of \affe is its principal type inference.
We now formulate our inference algorithm
based on the $HM(X)$ framework~\citep{DBLP:journals/tapos/OderskySW99}.
$HM(X)$ is a Hindley-Milner type system for a language
with qualified types where constraints are expressed in an arbitrary
theory $X$.
Most importantly,
$HM(X)$ ensures principal type inference
when $X$ respects the specified properties.
We apply HM(X) to a concrete constraint language which we name $\CL$.
We adapt and extend HM(X) existing rules to support kind inference,
track linearity and handle borrows and regions. Finally, we
formulate an appropriate constraint solving algorithm for $\CL$
which solves and simplifies constraints in a principled way.
We start by various preliminary definitions.

@@

The relations used for syntax-directed typing can trivially be defined
in term of these new relations by using constraint entailment.
All relations are fully described in \cref{typ:extra:envs}.

@@

\subsection{Kind inference}

We note $\inferK{(C,\unif)}{\bf{\E}}{\bf{\tau}}{k}$ when type $\tau$ has kind $k$
in environment $\E$ under constraints $C$ and unifier $\unif$.
$\E$ and $\tau$ are the input parameters of
our inference procedure.
We present the kind inference algorithm as a set of rules in
\cref{rules:kinding}.
Higher-kinds are not generally supported
and can only appear by looking-up the kind scheme of a type constructor,
for use in the type application rule {\sc KApp}.
Type variables must be of a simple kind in rule {\sc KVar}.
Kind schemes are instantiated in the {\sc KVar} rules by creating
fresh kind variables and the associated substitution.
{\sc KArr} and {\sc KBorrow} simply returns the kind of the primitive
arrow and borrow types.
The $\operatorname{normalize}$ function is used every time several constraints
must be composed in order to simplify the constraint and return a most general
unifier.

\begin{figure}[ht]
  \centering
  \input{infer/kinds}
  \caption{Kind inference algorithm -- $\inferK{(C,\unif)}{\E}{\tau}{k}$}
  \label{rules:kinding}
\end{figure}

@@

\subsubsection{Environments and bindings}
\label{infer:envs}

In our syntax-directed system, we ensured
that linear variables are not discarded at the \emph{leafs}, through
the {\sc Var} rule. In the inference algorithm, we operate in the opposite
direction: we collect data from the leafs, and only enforce linearity
at \emph{binders}. This is reflected in the {\sc Var$_I$} and
{\sc Abs$_I$} rules presented in \cref{rule:infer:envs}.
Typing for variables is very similar to traditional (non-affine) ML
typing. The only difference is that we collect
the fact that $x$ was used with the scheme $\schm$ by returning
a usage environment $\Sv = \{ \bvar{x}{\schm} \}$.
%
This usage environment is in turn used at each binder to enforce proper
usage of linear variable via the $\Weaken$ property.
We demonstrate how this is achieved for lambda expressions
in the {\sc Abs$_I$} rule.
First, we typecheck the body of the lambda and obtain a usage
environment $\Sv_x$. Just like in the syntax directed,
we introduce the constraint
$\Cleq{\Sv\Sdel{x}}{\kvar}$ which properly accounts for captures in
the body of the lambda expression. We also introduce the constraint
$\Weaken_{\bvar{x}{\sigma}}(\Sv)$. If $x$ was used $e$, $x$ is present
in $\Sv$ and this constraint is empty. Otherwise, if $x$ is
\emph{not} present in $\Sv$, its type should not have a linear
kind, which is enforced by the constraints $\Cleq{\sigma}{\kaff_\infty}$.
The $\Weaken$ constraint is introduce at each binding point in the program
such as let-bindings and pattern matching.

@@

\begin{figure}[!h]
  \begin{mathpar}
    { \Weaken_{\bvar{x}{\sigma}}(\Sv) =
      \begin{dcases}
        \Ctrue&\text{if } x\in\Sv\\
        \Cleq{\sigma}{\kaff_\infty}&\\
      \end{dcases}
    }
    \and
    \ruleIVar
    \and
    \ruleIAbs
  \end{mathpar}
  \caption{The {\sc Abs$_I$} and {\sc Var$_I$} rules}
  \label{rule:infer:envs}
\end{figure}

\subsubsection{Splitting and Regions}
\label{infer:split}
\label{infer:regions}

We now present the inference version
of the {\sc Pair} and {\sc Region} rules in \cref{rule:infer:envrules}.
While the rules are quite similar to the original version, the usage
environment $\Sv$ is now an \emph{output} of the algorithm.
As such, we use the ``inference'' version of the relations on
environment,
$\bsplit{C}{\Sv}{\bf{\Sv_1}}{\bf{\Sv_2}}$ and $\bregion{C}{\bf{x}}{\Sv}{\bf\Sv'}$,
which return the necessary constraints.
As appropriate for inference, these relations 
are total and non ambiguous for their input parameters ($\Sv_1, \Sv_2$ and $\Sv'$, respectively).
We then collect all constraints and normalize them.

\begin{figure}[!h]
  \centering
  \begin{mathpar}
    \ruleIPair

    \ruleIRegion
  \end{mathpar}
  \caption{The {\sc Pair$_I$} and {\sc Region$_I$} rules}
  \label{rule:infer:envrules}
\end{figure}

\subsubsection{Generalization and constraints}

So far, constraints have only been composed of a list of kind inequalities.
The internal language of constraints used for inference is in fact
richer.
To illustrate this, we show the inference rule for let-bindings
in \cref{rule:infer:let}.
This rule combines several ingredients previously seen:
since let expressions are binders, we use $\Weaken$ on the bound
identifier $x$. Since let-expressions contain
two subexpressions, we use the environment splitting relation,
$\bsplit{C_s}{\Sv}{\Sv_1}{(\Sv_2 \Sdel{x})}$. We remove the $x$ from
the right environment, since it is not available outside of the expression
$e_2$, and should not be part of the returned usage environment.

As per tradition in ML languages, generalization is performed
on let-bindings.
Following HM(X), we note $(C_\schm, \schm) = \generalize{C}{\E}{\tau}$
the pair of a constraint and a scheme resulting from
generalization. The definition is provided in \cref{rule:infer:let}.
The type scheme $\schm$ is created by quantifying over all the appropriate
free variables and the current constraints.
The generated constraint $C_\schm$ uses a new projection operator,
noted $\Cproj{\tvar}{D}$, which
allow the creation of local variables inside the constraints.
This allow to encapsulate all the quantified variables in the global constraints.
It also reflect the fact that there
should exists at least one solution for $C$ for the scheme to be valid.
\citet{DBLP:journals/tapos/OderskySW99} give a detailed account
on the projection operators in HM inference.

\begin{figure}[!h]
  \centering
  \begin{mathpar}
    \ruleILet

@@

      \Multi{\kvar},\Multi{\tvar} = (\fv{\tau}\cup\fv{C})\setminus\fv{\E}
    }
  \end{mathpar}
  \caption{The {\sc Let$_I$} rule}
  \label{rule:infer:let}
\end{figure}

@@

An example of this algorithm in action is shown in \cref{solving:example}.
Our algorithm is complete, computes principal normal forms,
and already simplifies constraints significantly
(thanks to steps 6, 7 and 8).
It can be extended with further simplification phases.
In particular, our implementation and all the signatures presented in
\cref{motivation} use a variance-based simplification
where all covariant (resp. contravariant) variables are replaced by their
lower (resp. upper) bounds.
All the simplification mechanisms presented
here, including the variance-based one, are complete.
It is also possible to add ``best-effort'' simplification
rules which help reduce the size of inferred signatures even further
\citep{DBLP:conf/aplas/Simonet03}.
%
We now show that this algorithms is well-behaved with respect to entailment
and computes unique normal forms.

@@

We now show that our extended inference algorithm still satisfies the soundness
and completeness properties of HM(X). This is achieved by extending
the original proofs from HM(X), which is done in \cref{appendix:infer}.
%
The first theorem states that our inference algorithm is sound
with respect to the syntax directed type system.

\begin{theorem}[Soundness of inference]
  Given a type environment $\E$ containing only value bindings,
  $\E|_\tau$ containing only type binding and a term $e$,\\
  if $\inferW{\Sv}{(C,\unif)}{\E;\E_\tau}{e}{\tau}$
  then $\inferS{C}{\unif(\Sv;\E_\tau)}{e}{\tau}$, $\unif C = C$ and $\unif \tau = \tau$
\end{theorem}

@@

The second theorem states that our algorithm is complete: for any given
syntax directed typing derivation, our inference algorithm can find
a derivation that gives a type at least as general.
For this, we need first to provide a few additional definitions.

>>>>

An important contribution of \affe is its principal type inference.
We now formulate our inference algorithm
based on the $HM(X)$ framework~\citep{DBLP:journals/tapos/OderskySW99}.
$HM(X)$ is a Hindley-Milner type system for a language
with qualified types where constraints are expressed in an arbitrary
theory $X$.
Most importantly,
$HM(X)$ ensures principal type inference
when $X$ respects the specified properties.
We apply HM(X) to a concrete constraint language which we name $\CL$.
We adapt and extend HM(X)'s existing rules to support kind inference,
track linearity and handle borrows and regions. Finally, we
formulate an appropriate constraint solving algorithm for $\CL$
which solves and simplifies constraints in a principled way.
We start by various preliminary definitions.

@@

The relations used for syntax-directed typing can trivially be defined
in terms of these new relations by using constraint entailment.
All relations are fully described in \cref{typ:extra:envs}.

@@

\subsection{Kind inference}
%
\begin{figure}[t]
  \centering
  \input{infer/kinds}
  \caption{Kind inference algorithm -- $\inferK{(C,\unif)}{\E}{\tau}{k}$}
  \label{rules:kinding}
\end{figure}
%
We note $\inferK{(C,\unif)}{\bf{\E}}{\bf{\tau}}{k}$ when type $\tau$ has kind $k$
in environment $\E$ under constraints $C$ and unifier $\unif$.
$\E$ and $\tau$ are the input parameters of
our inference procedure.
We present the kind inference algorithm as a set of rules in
\cref{rules:kinding}.
Higher-kinds are not generally supported
and can only appear by looking-up the kind scheme of a type constructor,
for use in the type application rule {\sc KApp}.
Type variables must be of a simple kind in rule {\sc KVar}.
Kind schemes are instantiated in the {\sc KVar} rules by creating
fresh kind variables and the associated substitution.
{\sc KArr} and {\sc KBorrow} simply returns the kind of the primitive
arrow and borrow types.
The $\operatorname{normalize}$ function is used every time several constraints
must be composed in order to simplify the constraint and return a most general
unifier.

@@

\subsubsection{Environments and bindings}
\label{infer:envs}
%
\begin{figure}[tp]
  \begin{mathpar}
    { \Weaken_{\bvar{x}{\sigma}}(\Sv) =
      \begin{dcases}
        \Ctrue&\text{if } x\in\Sv\\
        \Cleq{\sigma}{\kaff_\infty}&\\
      \end{dcases}
    }
    \and
    \ruleIVar
    \and
    \ruleIAbs
  \end{mathpar}
  \caption{The {\sc Abs$_I$} and {\sc Var$_I$} rules}
  \label{rule:infer:envs}
\end{figure}
%
In our syntax-directed system, we ensured
that linear variables are not discarded at the \emph{leaves}, through
the {\sc Var} rule. In the inference algorithm, we operate in the opposite
direction: we collect data from the leaves, and only enforce linearity
at \emph{binders}. This is reflected in the {\sc Var$_I$} and
{\sc Abs$_I$} rules presented in \cref{rule:infer:envs}.
Typing for variables is very similar to traditional (non-affine) ML
typing. The only difference is that we collect
the fact that $x$ was used with the scheme $\schm$ by returning
a usage environment $\Sv = \{ \bvar{x}{\schm} \}$.
%
This usage environment is in turn used at each binder to enforce proper
usage of linear variable via the $\Weaken$ property.
We demonstrate how this is achieved for lambda expressions
in the {\sc Abs$_I$} rule.
First, we typecheck the body of the lambda and obtain a usage
environment $\Sv_x$. Just like in the syntax-directed type system,
we introduce the constraint
$\Cleq{\Sv\Sdel{x}}{\kvar}$ which properly accounts for captures in
the body of the lambda expression. We also introduce the constraint
$\Weaken_{\bvar{x}{\sigma}}(\Sv)$. If $x$ was used $e$, $x$ is present
in $\Sv$ and this constraint is empty. Otherwise, if $x$ is
\emph{not} present in $\Sv$, its type should not have a linear
kind, which is enforced by the constraints $\Cleq{\sigma}{\kaff_\infty}$.
The $\Weaken$ constraint is introduced at each binding point in the program
such as let-bindings and pattern matching.

@@

\subsubsection{Splitting and Regions}
\label{infer:split}
\label{infer:regions}
\begin{figure}[tb]
  \centering
  \begin{mathpar}
    \ruleIPair

    \ruleIRegion
  \end{mathpar}
  \caption{The {\sc Pair$_I$} and {\sc Region$_I$} rules}
  \label{rule:infer:envrules}
\end{figure}
%
We now present the inference version
of the {\sc Pair} and {\sc Region} rules in \cref{rule:infer:envrules}.
While the rules are quite similar to the original version, the usage
environment $\Sv$ is now an \emph{output} of the algorithm.
As such, we use the ``inference'' version of the relations on
the environment,
$\bsplit{C}{\Sv}{\bf{\Sv_1}}{\bf{\Sv_2}}$ and $\bregion{C}{\bf{x}}{\Sv}{\bf\Sv'}$,
which returns the necessary constraints.
As appropriate for inference, these relations
are total and non ambiguous for their input parameters ($\Sv_1, \Sv_2$ and $\Sv'$, respectively).
We then collect all constraints and normalize them.

\subsubsection{Generalization and constraints}
%
\begin{figure}[tbp]
  \centering
  \begin{mathpar}
    \ruleILet

@@

      \Multi{\kvar},\Multi{\tvar} = (\fv{\tau}\cup\fv{C})\setminus\fv{\E}
    }
  \end{mathpar}
  \caption{The {\sc Let$_I$} rule}
  \label{rule:infer:let}
\end{figure}
%
So far, constraints have only been composed of a list of kind inequalities.
The internal language of constraints used for inference is in fact
richer.
To illustrate this, we show the inference rule for let-bindings
in \cref{rule:infer:let}.
This rule combines several ingredients previously seen:
since let expressions are binders, we use $\Weaken$ on the bound
identifier $x$. Since let-expressions contain
two subexpressions, we use the environment splitting relation,
$\bsplit{C_s}{\Sv}{\Sv_1}{(\Sv_2 \Sdel{x})}$. We remove the $x$ from
the right environment, since it is not available outside of the expression
$e_2$, and should not be part of the returned usage environment.

As per tradition in ML languages, generalization is performed
on let-bindings.
Following HM(X), we note $(C_\schm, \schm) = \generalize{C}{\E}{\tau}$
the pair of a constraint and a scheme resulting from
generalization. The definition is provided in \cref{rule:infer:let}.
The type scheme $\schm$ is created by quantifying over all the appropriate
free variables and the current constraints.
The generated constraint $C_\schm$ uses a new projection operator,
noted $\Cproj{\tvar}{D}$, which
allow the creation of local variables inside the constraints.
This allows us to encapsulate all the quantified variables in the global constraints.
It also reflects the fact that there
should exist at least one solution for $C$ for the scheme to be valid.
\citet{DBLP:journals/tapos/OderskySW99} give a detailed account
on the projection operators in HM inference.

@@

An example of this algorithm in action is shown in \cref{solving:example}.
Our algorithm is complete, computes principal normal forms,
and already simplifies constraints significantly
(thanks to steps 6, 7 and 8).
It can be extended with further simplification phases.
In particular, our implementation and all the signatures presented in
\cref{motivation} use a variance-based simplification
where all covariant (resp. contravariant) variables are replaced by their
lower (resp. upper) bounds.
All the simplification mechanisms presented
here, including the variance-based one, are complete.
It is also possible to add ``best-effort'' simplification
rules which help reduce the size of inferred signatures even further
\citep{DBLP:conf/aplas/Simonet03}.
%
We now show that this algorithm is well-behaved with respect to entailment
and computes unique normal forms.

@@

We now show that our extended inference algorithm still satisfies the soundness
and completeness properties of HM(X). This is achieved by extending
the original proofs from HM(X), which is done in \cref{appendix:infer}.
%
The first theorem states that our inference algorithm is sound
with respect to the syntax-directed type system.

\begin{theorem}[Soundness of inference]
  Given a type environment $\E$ containing only value bindings,
  $\E|_\tau$ containing only a type binding and a term $e$,\\
  if $\inferW{\Sv}{(C,\unif)}{\E;\E_\tau}{e}{\tau}$
  then $\inferS{C}{\unif(\Sv;\E_\tau)}{e}{\tau}$, $\unif C = C$ and $\unif \tau = \tau$
\end{theorem}

@@

The second theorem states that our algorithm is complete: for any given
syntax-directed typing derivation, our inference algorithm can find
a derivation that gives a type at least as general.
For this, we need first to provide a few additional definitions.
